On simplification and gravel rivers: Einstein and Bagnold in a modern context

### 2021-09-03 revision/formerPapers/com-piece/scraps/com-piece2.tex
added phd thesis.

\title{On argumentation and gravel rivers: Einstein and Bagnold in a modern context}
\author{Kevin Pierce}
\maketitle

%% intro

Gravel bed rivers are troubled natural systems, and changes in their morphology are linked to a wide array of ecological and engineering problems.
Since the transport of coarse material as bedload is an important control on channel morphology, bedload transport has been the subject of intense research for over a century. 
Despite this effort, our ability to predict bedload transport rates is inadequate.  
In natural streams, predictions of bedload transport can deviate from measured values by factors of 100 or 1000. 
Since the bedload transport process is governed by Newtonian mechanics, it should be fully characterized by exchanges of momentum between the granular and fluid phases, highlighting a mechanical description as an attractive prospect.
However, because bedload transport results from a turbulent flow interacting with granular matter, the full mechanical description is a fusion of two notoriously difficult and tenuously understood subfields of classical physics, meaning many of the required details are missing.
In absence of these details, many researchers have attempted to reverse engineer mechanical theories of bedload transport, and among these, two of them have been taken up by generations of researchers, forming two main research paradigms in the physics of bedload transport. 
These are the stochastic approach, initiated by \citet{Einstein1937, Einstein1950, Einstein1964} and the deterministic approach initiated by \citet{Bagnold1956, Bagnold1966, Bagnold1973}.

Each of these approaches is founded on the application of mechanical reasoning to a simplified concept of bedload motion, characterized by empirical parameters, so that both categories of models must be calibrated to applications. 
Also, the simplified concepts of Einstein and Bagnold-type models are different with regard to most aspects of bedload motion.
Accordingly, when proponents of different schools discuss sediment transport processes, they often find miscommunications and disagreements. 
This conflict is exascerbated by observations that neither mechanical approach is, practically speaking, very good; and that neither approach is clearly superior than the other. 
Specifically, both of these mechanical paradigms are expected to generate poorer bedload transport predictions than simpler power law curves, calibrated to empirical data without reference to mechanical concepts \citep{Barry2004}; and the semi-empirical aspect of these models provides flexibility, so comparing them by predictive ability is difficult, as neither approach shows uniform superiority across a full range of flow and sediment characteristics. 
Accordingly, although each mechanical approach has clear merits and demerits, and it's safe to conclude neither is perfect; the perspective that either is fully superior, for any reason apart from practicality, lacks objective support. 
Because there is no clear path to join the two schools together, proponents of each school toil away in their respective camps, solving problems within their own simplified conceptual pictures, and failing to communicate science across the stochastic / deterministic rift that Einstein and Bagnold have left behind. 
We need to ask: is this rift justified by evidence? Are Bagnold and Einstein really so different that they share no common ground, and is one or the other completely wrong? If so, theorists should out the failed approach and concentrate on its successor together. If not, we should work together to synthesize something new out of the strongest elements of each paradigm.

To answer these questions, we should carefully assess the concepts of each paradigm, and try to discern any conditions of bedload transport under which one description provides more or less useful information than the other. 
Because bedload transport rates are the cumulative result of individual particle motions, an appropriate starting point seems to be a clear physical picture of individual bedload motions, and this picture should be consistent with contemporary experiments. 
Bedload grains typically move intermittently, rolling, sliding, and bouncing along the river bed \citep{Einstein1950, Bagnold1973}.
% resting, entrainment by impulse, motion in turbulent flow, collisions with bed, deposition

@@

% first describe broad brush ideas. initiation, motion, cessation 
To approach the question, we should develop a clear mechanical understanding of individual bedload motions which is supported by experiments, and we should consider where Einstein and Bagnold's theories fall in relation to it.  
Because bedload motion is intermittent, a common and useful concept is to divide individual particle motions into three stages: initiation, movement, and cessation. 
To interpret bedload experiments, investigators have studied how the properties of each stage result from a balance of granular and fluid interactions in accord with Newtonian mechanics. 
Initiation, or the transition of a bed particle from rest to motion, is undoubtedly a result of momentum transfer to the stationary particle, by the fluid and possibly by impacts from other moving grains, and these driving impulses are resisted by frictional and contact forces imparted on stationary grains by every other grain in contact. 
Similarly, the trajectories of particles, once they are in movement, are determined by the momentum transferred from the fluid phase and by impulses imparted by collisions with the irregular bed surface formed by stationary grains \citep{Wiberg1985, Bialik2015}. 
The specifics of the fluid and granular forces are not accessible: to fully characterize the fluid forces would require the full time series of pressure and velocity fields within the turbulent fluid, including within the porous bed, and to characterize the static forces and dynamic impulses imparted by the bed surface would require knowledge of the location and relative orientation of every grain within the bed. 
Accordingly, the initiation and trajectories of particles is very difficult to understand, but the cessasion of motion is even more difficult. 
Cessasion of motion, or the transition of a moving particle to rest, results from a chance coordination of upstream impact forces from the configuration of static grains with a lull in turbulent driving forces \citep{Pahtz2018}, again in accord with Newton's laws.

>>>>

\title{On simplification and gravel rivers: Einstein and Bagnold in a modern context}
\author{Kevin Pierce}
\maketitle

\section{Introduction: Gravel bed rivers and mechanistic theories of bedload motion}
%% gravel bed rivers, bedload transport, mechanistic theories, Einstein and Bagnold 
Gravel bed rivers are troubled natural systems, and changes in their morphology are linked to a wide array of ecological and engineering problems.
Since the transport of coarse material as bedload is an important control on channel morphology, bedload transport has been the subject of intense research for over a century. 
Despite this effort, our ability to predict bedload transport rates is inadequate.  
In natural streams, predictions of bedload transport can deviate from measured values by factors of 100 or 1000. 
Prediction is difficult because individual bedload grains move intermittently, rolling, sliding, and bouncing along the river bed \citep{Einstein1950, Bagnold1973}. 
Since this bedload transport process is governed by Newtonian mechanics, it should be fully characterized by exchanges of momentum between the granular and fluid phases, highlighting a mechanical description as an attractive prospect.
However, because bedload transport results from a turbulent flow interacting with granular matter, the full mechanical description is a fusion of two notoriously difficult and tenuously understood subfields of classical physics, meaning many of the required details are missing.
In absence of these details, many researchers have attempted to reverse engineer mechanical theories of bedload transport, and from these efforts, two of them have emerged as important research paradigms in the physics of bedload transport. 
These are the stochastic approach, initiated by \citet{Einstein1937, Einstein1950, Einstein1964} and the deterministic approach, initiated by \citet{Bagnold1956, Bagnold1966, Bagnold1973}.

% Einstein and Bagnold -- simplified concepts of motion -- semi-empirical -- rift between the two schools -- lack of communication between them 
These two paradigms are founded on the application of mechanical reasoning to simplified concepts of bedload motion.
These concepts are characterized by empirical parameters, so that mechanical models must be calibrated to applications. 
Owing to these simplified concepts, proponents of Einstein and Bagnold-type models have different understanding of most aspects of bedload transport.
Accordingly, when proponents of different schools discuss sediment transport processes, they often find miscommunications and disagreements. 
This conflict is exascerbated by observations that neither mechanical approach is, practically speaking, very good; and any superiority of either approach is difficult to judge. 
For example, either mechanical paradigm generates poorer bedload transport predictions than simpler power law curves, calibrated to empirical data without reference to mechanical concepts \citep{Barry2004}; and the semi-empirical aspect of these models provides flexibility, so comparing them by predictive ability is difficult, as neither approach shows uniform superiority across a full range of flow and sediment characteristics. 
Accordingly, although each mechanical approach has merits, a perspective that either is uniformly superior for any reason apart from practicality, is not supported by evidence.  
Because there is no clear path to join the two paradigms together, proponents of each toil away in their respective camps, solving problems within their own simplified conceptual pictures, and failing to communicate science across the stochastic / deterministic rift. 
We need to ask: is this scientific divide justified by evidence? 
Are stochastic and deterministic approaches to understand bedload transport as a mechanical process really so different that they share no common ground, and is one or the other completely wrong? 
If so, theorists should out the failed approach and focus together on developing its successor. 
If not, we should work together to synthesize something new out of the strongest elements of each paradigm.

\section{Bedload fluxes from individual motions: the stochastic paradigm of Einstein}
% Einstein -- bedload flux is individual motions -- individual motions are random -- start to stop transitions -- entrainment, motion, and deposition

Einstein initiated the stochastic paradigm by relating bedload transport rates to individual particle motions \citep{Einstein1942, Einstein1950}.
He conceptualized the downstream transport of an individual particle as an alternating series of rests and hops, and he set out to determine the period of rest and distance covered during a hop with flume experiments \citep{Einstein1937}. 
During these experiments, he realized that the resting period and hop distance of individual particles was not consistent or predictable from one particle to the next, so he was forced to switch to a statistical conception. 
Viewing these motion characteristics on a population level involving many particles, Einstein noted the resting period and hop distance of individual particles lie on stable statistical distributions, which he measured \citep{Einstein1937}. 
Initially, he used the random characteristics of individual rests and motions to calculate the probability distribution of the effective velocity at which bedload grains travel as the move downstream through cycles of rest and motion, with relevance to the spreading or diffusion of tracer particles as they move downstream due to their different effective velocity characteristics \citep{Einstein1937}. 
Later, he derived the random character of the resting period as a mechanical consequence of fluid turbulence, and he used this concept to derive the mean rate of bedload transport \citep{Einstein1950}. 
The tenacity with which Einstein's ideas have been followed up, modified, critiqued, and improved over the last seventy years is a testament to his pioneering originality \citep{Ettema2004}, yet the clear pattern of development and refinement along the course of his life provides motivation, showing an obvious balance between hard work and genius \citep{Einstein1937, Einstein1942, Einstein1950, Einstein1964}.

Generations of researchers and proponents of Einstein have criticized, generalized, and reinterpreted his stochastic concept of bedload transport, without fully invalidating his ideas. 
Statistical characterizations of individual motions have been investigated extensively \citep{Hubbell1964, Yano1969, Nakagawa1976, Hassan1991, Habersack2001, Ancey2008, Roseberry2012, Heyman2016}, and it's clear that his conception of their characteristics as random variables drawn from statistical distributions is essentially correct. 
Since Einstein, new characteristics of particle motion have been considered, and incorporated into stochastic theories of bedload transport, such as the velocity, acceleration, and travel time during single particle motions \citep{Drake1988, Ancey2008, Lajeunesse2010, Roseberry2012, Fathel2015, Heyman2016}. 
Understanding of particle diffusion has advanced a great deal, and the problem is obviously more nuanced than Einstein considered \citep{Einstein1937}.  
Particle diffusion has different characteristics depending on the temporal or spatial scales of consideration \citep{Nikora2002, Zhang2012, Martin2012}, and this probably relates crucially to the effects of particle burial \citep{Sayre1971, Nakagawa1980, Voepel2013, Martin2014, Bradley2017} and variable movement characteristics due to sediment size \citep{Fan2017} on the differences in effective velocities of moving bedload.

Most notably, Einstein's mechanical theory of the bedload transport rate \citep{Einstein1950} has been reworked a great deal. 
Subsequent theories on the resting time of stationary sediment, which is equivalent to the rate of erosion of an individual surface grain, took account of the randomness in supporting configuration of bed surface grains, in addition to randomness resulting from fluid turbulence \citep{Paintal1971, Wu2002, Dey2018}, and experiments have increasing recognition of the role of both the magnitude and duration of turbulent fluctuations on particle entrainment \citep{Diplas2008, Valyrakis2010, Celik2014}, which Einstein-like models of the resting time (or erosion rate) have not yet incorporated \citep{Dey2018}. 
Additionally, instead of computing mean bedload rates only, contemporary theories have harnessed the physics of stochastic processes to compute probability distributions of the bedload flux, so that mean values and the expected magnitude of bedload fluctuations are available \citep{Lisle1998, Sun2000, Ancey2006, Ancey2008, Ma2014, Ancey2014}. 
This work reflected back on the resting period of particle motions in a contentious way which still requires resolution. 
Einstein's concepts, when translated to a theory which predicts the magnitude of bedload fluctuations, produces fluctuations which are too small to describe experiments \citep{Ancey2006}. 
Accordingly, a set of contemporary models has incorporated collective feedback mechanisms in particle motion, which could be physically attributed to the erosion of multiple grains from the bed at once \citep{Ancey2008, Heyman2013, Ma2014}, as might arise from small granular avalanches \citep{Heyman2013}, the effects of coherent turbulence on particle motion \citep{Nino1998, Amir2014, Santos2014, Shih2017}, or the migration of bedforms \citep{Dhont2018}. 
In any case, research from the stochastic paradigm which Einstein initiated has generated a formidable amount of understanding about the physics underlying bedload transport, and this progress is expected to continue.

However, many elements of Einstein's conception can be criticized. 
For one thing, there are multiple modes within the statistics characterizing particle rest and motion, and no clear indiciation of the extent to which these must be resolved in bedload transport theory. 
For example, particles can rest on the bed surface, and this resting period can be characterized probabilistically \citep{Einstein1950}, but these resting times are known to be variable across bedforms, when the relative exposure of bed particles to the flow is variable \citep{Crickmore1962, Sayre1967, Yang1971, Nakagawa1980}. 
Also, particles can rest within the bed surface, or in a state of burial \citep{Voepel2013, Martin2014, Olinde2015, Bradley2017}, and the resting statistics associated with the burial process are different than those associated with surface resting. 
Similarly, when bedload particles move, they can fall into different modes of motion. 
For example, they can roll, slide, or bounce, and within each of these modes they may have different motion statistics \citep{Frey2006, Frey2014}. 
The degree to which these different modes must be accounted for in stochastic bedload transport theory has not been clarified.  
For another thing, at high transport stages, Einstein's partition of bedload transport into separate motion and rest periods breaks down.
Under these conditions, bedload motion takes the form of a sheet flow separating the bed from the water flow, and the entire bed surface moves with prevalent interactions between all moving grains \cite{Jenkins1998, Clark2015}, possibly with a vertical velocity distribution which depends upon distance from the shearing fluid \citep{Bagnold1973, Frey2011, Mouilleron2009, Houssais2015, Ferdowski2016}.
This stage of bedload transport is expected to require a different description.

\section{Bedload fluxes from collective dissipation: the deterministic paradigm of Bagnold}

%"According to the bagnold hypothesis, the fluid shear stress at the bed equals the threshold value corresponding to incipient motion of particles; the difference %between the total bottom shear stress provides the shear stress in the solid phase. (Ancey2006) Interpreting the excess stress as a result of momentum transfers %between the solid and fluid phases makes it possible to compute the number of particles that can be entrained and maintained in motion, and thereby the solid flow %rate"

Bagnold formed a different simplified concept of bedload transport to understand it with mechanical principles, initatiating the deterministic paradigm of mechanistic bedload transport theory. 
From the physical point of view, Bagnold considered bedload transport in terms of a momentum exchange between solid and liquid phases.
He balanced the average power supplied to the flow by gravity against the average frictional dissipation by moving bedload, and neglected any details of turbulent fluctuations, irregularity in the configuration of stationary grains, or the trajectories of moving particles.
Bagnold's formulation hinges on the application of an average dynamic friction coefficient to moving bedload in order to characterize its energy dissipation. 
His formulation is built upon a key hypothesis: when bedload transport is at its equilibrium value, so it dissipates all available power to move sediment, the tractive force imparted by the moving fluid on the bed surface will match to the threshold for sediment motion. 
This is the so-called Bagnold hypothesis \citep{Engelund1976, Luque1976, Seminara2002, Ancey2006}.
Bagnold's original formulation predicts a bedload transport rate which scales, at high stages of transport, as the average fluid shear stress to the 3/2 power: $q_b \propto \tau^{3/2},$ and his hypothesis has been taken up by generations of researchers, an enduring legacy which is even more interesting in that his bedload transport research seems to be at most a tertiary component of illustrious life's work, with a larger portion of his time devoted to his career as a decorated military commander in both world wars, and his initially recreational hobby of driving cars around the Sahara desert, which eventually gained him status as a preminent desert explorer \citep{Bagnold1988}.

Bagnold's original concept is enlightening for its pragmatic approach \citep{Ashida1972, Bagnold1973, Engelund1976, Luque1976}: his deterministic picture of bedload motion is simplified just enough from the real complexity to make contact with a foundational and reliable concept of classical physics, the conservation of energy. 
At the same time, his highly simplified assumptions and comparisons between experiements and his results have left many open questions for proponents of his ideas.

% Many investigators have built upon and refined Bagnold's concepts while preserving his main idea that bedload transport rates as a consequence of an average energy budget. 
% Early on, investigators came to the conclusion that the details of bedload trajectories are not negligible in a Bagnold-like formulation, and that bedload trajectories cannot be understood without accounting for turbulent lift forces.  
% suggesting a contingence of the bagnold formulation on the details of particle motions ( Luque and Van Beek 1976 )
% For one thing, lifting forces on sediment grains, which are highly variable in accord with fluid turbulence, contribute to the relative amount of time that bedload grains spend in contact with the bed, meaning the dynamic friction coefficient bagnold introduced should scale with fluid lift. 
% accordingly, a large set of bagnold-type treatments focused on including fluid turbulence in-so-far as it affects bedload trajectories, but not in relation to the average power balance which controls, in Bagnold's view, the average number of partilces in motion. 
% These studies effectively contend that bedload is dominated by an average power balance, but the dissipation mechanism within this balance is contingent on fluid turbulence.

For example, Bagnold's total neglect of fluid turbulence and any irregularity between sediment grains left open problems around the degree to which fluid turbulence effects and the differential mobility of grains of varying size can be incorporated into a Bagnold-type framework.  
Likewise, Bagnold's agnostic approach to individual bedload motions within his theory led many investigators to consider how these details can be reconciled with his energy dissipation concept. 
Despite these simplifications, Bagnold's original theory is aptly descriptive of bedload transport under particular conditions: at high transport stages, bedload transport rates often appear to scale as the average shear stress to the $3/2$ power ($q_b \propto \tau^{3/2}$) in accord with his mechanical prediction, at least over flat beds sheared by a steady uniform flow \citep{MeyerPeter1948,Gomez1989,Schmeeckle2014, Elghannay2017}. 
However, Bagnold's theory is troubled at relatively low transport stages and arbitrarily sloping beds.
To describe bedload rates at low transport stages, the empirical components of Bagnold's theory must be set to unphysical values \citep{Engelund1976, Luque1976,  Nelson1995, Nino1998}; and his deterministic theory neglects a key attribute of bedload transport rates at low stages, namely that transport is highly intermittent, so that bedload transport rates exhibit strong fluctuations which Bagnold's theory does not constrain \citep{Bohm2004, Ancey2008}. 
Over arbitrarily sloping beds, there are theoretical arguments that Bagnold's hypothesis should flatly fail, even for relatively small bed slopes transverse to the flow direction \citep{Seminara2002}. 
These problems have motivated intense research within the Bagnold paradigm.

Many researchers have concentrated on computing individual particle trajectories from Newtonian physics, and incorporating these details with Bagnold's average energy dissipation concepts, which should fix the number of particles in motion \citep{Abbott1977, Bridge1984, Wiberg1989, Bridge1992, Nino1998, Bialik2015}. 
Because the details of particle trajectories are highly contingent on fluid turbulence and collisions with the irregular bed surface, this incorporation of bedload trajectories can be viewed as a kind of first order inclusion of turbulent forcing and the random arrangement of bed grains \citep{}, which are key elements of Einstein-type models. 
A difficult attribute of this problem is that energy dissipation due to bedload should happen primarily when bedload is in contact with the bed surface, but the relative fraction of time bedload spends in contact with the bed surface is highly contingent on fluid turbulence, and especially fluid lift, and Bagnold neglected both of these aspects. 
Hence, Bagnold's dynamic friction coefficient should not really be a constant across all transport stages, and is instead expected to depend upon the details of bedload trajectories, which are inherently statistical and incompletely understood \citep{Bialik2015}. 
For one thing, the details of forces imparted by a turbulent flow on bedload grains are incompletely understood \citep{Maxey1986, Schmeeckle2007, Dwivedi2010, Dwivedi2011}, with relatively important terms in the force balance aside from the typical drag and lift \citep{Bialik2015}.
These models of individual bedload motions remain somewhat troubled, usually overpredicting the velocity at which bedload grains travel \citep{Bialik2015}, although the research is steadily generating new understanding.

critical shear stresses have been difficult to consistently define and calculate
and there are indications of difficulty in defining the concept as a consistent consequence of underlying physics

Another deficiency of bagnold's concepts is the failure of the approach over arbitrarily sloping beds
initially, the observation that bagnold's hypothesis did not lead to physical conclusions even for a relatively negligible transverse tilt in the bed surface
motivated the development of an alternate formulation which preserved his average energy balance concept while understanding the number of partilces in motion as a balance of erosion and deposition rates
Parker2003 ammended the bagnold hypothesis by replacing it with an Einstein type hypothesis regarding a balance of erosion and deposition rates 
they assumed relationships between the entrainment and deposition rates and the excess shear stress which, again, do not agree with 
Charru 2004 approached a bagnold-type description with a balance of erosion and deposition rates of particles
which was extended in many ways by the Lajenesse tracer paper and the Charru2006
However it did not make explicit reference to bagnold concepts, and is somehwat of a parallel development with similar conclusions

There is an overall conclusion that Bagnold's concepts work better when many particles are in motion, and when particles are relatively small. 
In full mobility bedload (stage 3 transport), there is a continuous bedload layer with several mobile grain layers beneath the surface (Frey2011, Moullerion2009, Jenkins1998, Hsu2004, ).
Under these conditions, bedload motion takes the form of a granular creep, somewhat closer to the concepts of the earliest bedload transport investigations
"This condition, which corresponds with the classical notion of bedload transport as a ‘tractive carpet flow’ (DuBoys, 1879; Sohn, 1997), is typical of sand beds and can also occur in gravel beds under strong flows. Much of the experimental work on bedload transport has been conducted in this condition"
Jenkins1998 We focus on sheet flows in which the particles are supported by their collisional interactions rather than by the velocity fluctuations of the turbulent fluid and introduce what we believe to be the simplest theory for the collisional regime that captures its essential features.
Hsu2004 computes bedload transport in a regime where collision interactions between moving grains are the only upward force, considered dominant to turbulent fluctuations
Balance equations for the average mass, momentum and energy for the two phases are phrased in terms of concentration–weighted (Favre averaged) velocities
Mouilleron2009 develops a revised bagnold model based upon a uniform moving particle concentration in order to compute the vertical velocity distribution within a granular sheet flow 
Once a sheet-layer flow is fully developed, the simple exchanges assumed here may be insufficient to describe reality, as moving particles would be continuously interacting with each others. In this limit, continuous models based on energy dissipation inside the bed load layer [Bagnold, 1973] might be more adequate.
Engelund says "Hence the bed load forms a ))protection shield(( at higher bed load concentrations, which controls the erosion rate."

%PROBLEMS AND OBSERVATIONS:
%bagnold entirely ignores the details of individual motions. There have been many attempts to clarify the Newtonian physics underlying bedload trajectories, and to incorporate these details into a bagnold type framework where an average energy budget fixes the number of particles in motion under bedload equilibrium. An interesting issue is that the bulk friction coefficient should depend on the amount of time that bedload grains spend in suspension between successive bed impacts, which may in turn depend on the details of bed trajectories. Usually, velocities of bedload are overestimated, which may relate to the ommision of collisons between moving particles Lee2002, or the need to account for shear stress imparted on the bed by saltating particles Nino1998,Bialik2015

%bagnold is inadequate at low discharges VanBeek1976, Engelund1976, Nelson1995, where the solid transport rate is not suitably large to dissipate the excess energy available to move bedload grains Engelund1976, Luque1976, implying the bulk particle friction coefficient must be fit to unphysical values, higher than the bulk friction angle of sediment. for example, Nino1998 found a dynamic friction angle of 56.6 degrees was required to describe their experiments at low transport stage, i.e., well above the angle of repose for any natural sediment

% Ancey2008 found that, at low sediment transport stages, up to 90% of the excess energy was dissipated by fluid turbulence, while at relatively high transport stages, as much as 75 percent was dissipated by bedload transport, implying a transition between bagnold and einstein-type regimes

%The bagnold equations show no loss of linear stability in the coupled-fluid granular system so it fails to capture the physics of bedforms Ancey2006

%Bagnold does not adapt to disequilibrium bed load transport or to mixtures of particle sizes Seminara2002,although Bridge1992 attempted to extend it to particle sizes, there is good discussion in Seminara about how this is kind of ridiculous

%bagnold fails away from steady uniform flow conditions and especially over arbitrarily sloping beds Seminara2002. The latter problem can be rectified by appealing to Einstein-like concepts of Entrainment and depostion rates, but the former problem is not currently solvable

%The mean-field treatment applied by Bagnold conflicts with a noticeable feature of sediment transport: its intermittent nature at low flow rates. Ancey2006

%very difficult to clarify incipient motion conditions Mongomery1997 or bulk friction angles Kirchener1990 experimentally or theoretically, also there are indications that the optimal value of tau_c to use in bagnold type models is distinct from that which is theoretically obtained to characterize the incipient motion of sediment-- "lso, this formulation appears to be very dependent on the value of t?c, which in general is di?cult to estimate because it does not necessarily correspond to the value associated with the threshold of particle motion in the absence of bedload transport." Nino1998

% Charru2004 attempts to rectify a bagnold framework with empirical data under viscous flows. They find the areal density of moving particles varies with time, which they link to granular compaction processes. Later on, this framework which borrows bagnoldean concepts with Einsteinian concepts in accord with the Parker2003 extension, is developed to take account of tracer diffusion and to describe the ripple spacing on a sand bed.

% very difficult to extend bagnold concepts to multiple grain sizes: BridgeBennett1992 noted that different grains sizes have different mobilities on the bed surface, and attempted to extend the formulation to multiple grain sizes but implying that the fluid shear stress is reduced to one critical value above the big grains and another, different value over the small grains places a strain on what turbulent flows are capable of doing. seminara 2002

%Ancey2008 analyzed energy dissipation by turbulence and by momentum transfers with moving bedload they found at low solid discharges as much as 90 percent of the energy supplied by gravity was disspated by turbulence while at higher discharges, as much as 75 percent can be disspated by momentum transfers with moving bedload so we see a continuous transition between the Einstein regime, in which turbulence is the dominant dissipative mechanism of the fluid flow, and the Bagnold regime, in which friction of moving bedload becomes the dominant mechanism.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Similarity and Difference between Einstein and Bagnold: toward a collaborative synthesis}

% Similarity and Difference -- transport stages -- properties varies across transport stage; 
"Transport of gravels and larger materials (which always move as bedload except in very exceptional circumstances) has been divided into three stages (Jackson and Beschta, 1982; Ash- worth and Ferguson, 1989; Warburton, 1992): (1) finer mate- rials overpassing a locally static bed: the material resident on the bed does not participate in the transport – the mobile material originates upstream; (2) partial transport of local bed material: at any one time part of the bed remains static, but any grain might eventually move; (3) general motion of the grains on the bed: all grains are equally apt to be moved."
In light of our discussion of Einstein and Bagnold, we can conclude that each simplified conceptual picture really pertains to a specific stage of bedload motion. 
In either case, individual bedload trajectories are undoubtedly stochastic: whether they are considered in relation to an Einstein or Bagnold-type formulation. 
The essential difference between the two formulations is that, within Einstein-type formulations, the bedload flux is a collective result of individual bedload motions, while in Bagnold-type formulations, the individual motions are secondary to an over-arching collective limitation on the number of moving particles. 
Because within the Einstein framework the intermittency of individual motion is dominant,

\section{
% Synthesis -- Einstein and Bagnold are limit cases, but think like Wolman: both are really missing the real complexity -- the intermediate region
% so it's not surprising that they both fail -- intermediate region means feedbacks with morphology introduce interactions from higher scales
% there is a clear separation between the scales of transport and morphology evolution, meaning the feedback is accentuated
% potential critical phenomena and a mixture of scales driving concordant evolution of bedload transport and morphology
% implying a potentially unsteady transport rate
% the next century should be spent exploring these morphology-transport feedbacks in accord with the ideas of classical morphology

\section{Conclusion}

% Mechanistic descriptions are attractive because they provide a possibility to leverage physical principles to improve our understanding of bedload motion
% Owing to the complexity of fluid turbulence and granular physics, two main research paradigms have emerged, based upon simplified concepts -- Einstein and Bagnold
% Einstein is based on the random characteristics of discrete particle motions, and describes a bedload rate with stochastic characteristics, where the intermittency of individual particle motions is emphasized, and the rate is better characterized by its statistics than by its mean trend 
% Bagnold is based on the average characteristics of a population of moving particles, and describes a deterministic bedload rate which is fully characterized by the average characteristics of the granular phase and fluid flow. 
% I have the perspective that the full complexity of bedload transport simplifies approximately to the simplified concepts of Einstein in Bagnold in different limit cases
% Einstein pertains to low transport stages, when not many particles are moving during an observation period, and the role of fluid turbulence and irregularities in bed configuration are emphasized over the average characteristics of the granular and fluid phases, implying a highly fluctuating transport rate dominated by unsteady aspects of the driving and resisting forces
% Bagnold pertains to high transport stages, when relatively many particles are moving, so there is an effective population averaging over the instantaneous characteristics of individual motions, meaning randomness can cancel out, implying a mostly steady transport rate dominated by the steady aspects of the driving and resisting forces. 
% However, neither of these limit cases is ever fully realized in gravel bed rivers. 
% There is a need to synthesize these two limit cases from a more sophisticated framework which includes especially the interaction between bed morphology, which is formed by bedload transport and 
% To make this synthesis, we have to get past our partisan and uncollaborative nature, and realize these mechanical theories for what they are, and quit judging them as right or wrong: they are simplified conceptual pictures of reality, with differing degrees of accord to the real world, depending on the conditions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To attempt to answer these questions, we should start from a clear physical picture of bedload transport rates, and then consider where Einstein and Bagnold's theories fall in relation to it. 
Because bedload transport rates are the cumulative result of individual particle motions, an appropriate starting point seems to be a clear physical understanding of individual bedload motions. 
Because bedload motion is intermittent, a common and useful concept is to divide individual particle motions into three stages: initiation, movement, and cessation. 
To interpret bedload experiments, investigators have studied how the properties of each stage result from a balance of granular and fluid interactions in accord with Newtonian mechanics. 
Initiation, or the transition of a bed particle from rest to motion, is undoubtedly a result of momentum transfer to the stationary particle, by the fluid and possibly by impacts from other moving grains, and these driving impulses are resisted by frictional and contact forces imparted on stationary grains by every other grain in contact. 
Similarly, the trajectories of particles, once they are in movement, are determined by the momentum transferred from the fluid phase and by impulses imparted by collisions with the irregular bed surface formed by stationary grains \citep{Wiberg1985, Bialik2015}. 
The specifics of the fluid and granular forces are not accessible: to fully characterize the fluid forces would require the full time series of pressure and velocity fields within the turbulent fluid, including within the porous bed, and to characterize the static forces and dynamic impulses imparted by the bed surface would require knowledge of the location and relative orientation of every grain within the bed. 
Accordingly, the initiation and trajectories of particles is very difficult to understand, but the cessasion of motion is even more difficult. 
Cessasion of motion, or the transition of a moving particle to rest, results from a chance coordination of upstream impact forces from the configuration of static grains with a lull in turbulent driving forces \citep{Pahtz2018}, again in accord with Newton's laws.

% resting, entrainment by impulse, motion in turbulent flow, collisions with bed, deposition

@@

% first describe broad brush ideas. initiation, motion, cessation
